Finally Tagless, Partially Evaluated

### 2008-09-12 tagless/tagless-final.tex
Minor changes to make notation more uniform

% see tagless\_interp1.ml, module Tagfull for the complete code.
% If you change the code in here, please adjust the .ml file
% accordingly. Let the paper and the accompanying code be in sync.
It is straightforward to create an algebraic data type, say in OCaml%
\ifshort, Fig.~\ref{fig:tag-problem}(a)\fi, to
represent object terms such as those in Figure~\ref{fig:object}.
For brevity, we elide treating integers, conditionals, and fixpoint in
this section.
\ifshort\else\UseVerbatim{2a}\fi
We represent each variable using a unary de Bruijn index.%
\footnote{We use de Bruijn indices to simplify the comparison with
\citearound{'s work}\citet{WalidICFP02}.}
For example, we represent the object term $(\fun{x}x)\True$ as
\ifshort \BUseVerbatim{test1}.\else \UseVerbatim{test1}\fi
\ifshort

@@

\fi
Following \citet{WalidICFP02},
we try to implement an interpreter function |eval0|\ifshort,
Fig.~\ref{fig:tag-problem}(b)\fi. It takes
an object term such as |test1| above and gives us its value.
The first argument to |eval0| is the environment, initially empty,
which is the list of values bound to free variables in the
interpreted code.
\ifshort\else\UseVerbatim{2b}\fi
If our OCaml-like metalanguage were untyped, the code above would be 
acceptable.
The |L e| line exhibits interpretive overhead:
|eval0| traverses the function body~|e| every time (the result of
evaluating) |L e| is applied. Staging can be used to remove this
interpretive overhead \citep[\S1.1--2]{WalidICFP02}.

@@

The usual solution is to introduce a universal type \citep[\S1.3]
{WalidICFP02} containing both booleans and functions\ifshort,
Fig.~\ref{fig:tag-problem}(c)\fi.
\ifshort\else\UseVerbatim{2c}\fi
We can then write a typed interpreter\ifshort,
Fig.~\ref{fig:tag-problem}(d), \else\UseVerbatim{2d}\fi
whose inferred type is |u list -> exp -> u|. Now we can evaluate
\ifshort
\texttt{eval [] test1} obtaining |UB true|.
\else
\begin{code}
let test1r = eval [] test1
val test1r : u = UB true 
\end{code}
\fi
The unfortunate tag |UB| in the result reflects that |eval| is a partial
function.  
First, the pattern match |with UA f| in the line
|A (e1,e2)| is not exhaustive, so |eval| can fail if we apply a boolean,
as in the ill-typed term |A (B true, B false)|.
\ifshort\else
\begin{code}
let test2 = A (B true, B false)
let test2r = eval [] test2
Exception: Match_failure in eval
\end{code}
\fi
Second, the |lookup|
function assumes a nonempty environment, so |eval| can fail if we
evaluate an open term
\ifshort
|A (L (V (VS VZ)), B true)|.
\else
\begin{code}
let test3 = A (L (V (VS VZ)), B true)
let test3r = eval [] test3
Exception: Match_failure in lookup
\end{code}
\fi
After all, the type |exp| represents object
terms both well-typed and ill-typed, both open and closed.

@@

The evaluator above is wired directly into
functions such as |b|, |lam|, and |app|, whose names appear
free in |testf1| above.
\ifshort We \else In the rest of this paper, we \fi
explain how to abstract over these functions' definitions
and apply different folds to the same object language, so as
to process the same term using many other
interpreters: we can
\begin{itemize}
    \item evaluate the term to a value in the metalanguage;
    \item measure the size or depth of the term;
    \item compile the term, with staging support such as in MetaOCaml;
    \item partially evaluate the term, online; and
    \item transform the term to continuation\hyp passing style (CPS),
        even call-by-name (CBN) CPS in a call-by-value (CBV) metalanguage,
        so as to isolate the evaluation
        order of the object language from that of the metalanguage.\ifshort
\footnote{Due to serious lack of space, 
we refer the reader to the accompanying code for this.}\fi
\end{itemize}
We have programmed all our interpreters and examples in OCaml (and, for staging,
\citet{metaocaml}) and standard Haskell. The complete code is
available at \url{http://okmij.org/ftp/packages/tagless-final.tar.gz}
to supplement the paper. 
\ifshort For simplicity, main examples in the paper will be in MetaOCaml;
all examples have also been implemented in Haskell.
\else Except for the basic definitions in \S\ref{encoding},
we show our examples in (Meta)OCaml even though some of our
claims are more obvious in Haskell, for consistency and because
MetaOCaml provides convenient, typed staging facilities.
\fi

@@

Figure~\ref{fig:object} shows our object language, a simply-typed
$\lambda$-calculus with fixpoint, integers, booleans, and comparison.
The language is close to \citets{xi-guarded}, without their polymorphic
lift but with more constants so as to more conveniently express Fibonacci,
factorial, and power.
In contrast to \S\ref{intro}, in the rest of the paper 
we use higher-order
abstract syntax (HOAS) \citep{miller-manipulating,pfenning-higher-order}
rather than de Bruijn indices to encode binding and
ensures that our object programs are closed.
We find HOAS to be more convenient, but we have also implemented our
approach using de Bruijn indices
(in Section~\ref{S:de-Bruijn} and the accompanying source file
|incope-dB.ml|).

@@

We embed our language in (Meta)OCaml and Haskell.  In Haskell,
the functions that construct object terms are methods in a type class
|Symantics| (with a parameter |repr| of kind |* -> *|)\ifshort,
Fig.~\ref{fig:symantics-haskell}(a)\fi. The class is so named
because its interface gives the syntax of the object language and its
instances give the semantics.
\ifshort\else\UseVerbatim{3a}\fi
For example, we encode the term |test1|, or $(\fun{x}x)\True$, from
\S\ref{tagproblem} above as \texttt{app (lam (\textbackslash x -> x)) (bool True)},
whose inferred type is \texttt{Symantics repr => repr Bool}.
For another example, the classical $\mathit{power}$ function is
\ifshort in Fig.~\ref{fig:symantics-haskell}(b)
\else\UseVerbatim{3b}\fi
and the partial application $\fun{x} \mathit{power}\;x\;7$ is
\ifshort in Fig.~\ref{fig:symantics-haskell}(c).
\else\UseVerbatim{3c}\fi
The dummy argument |()| above is to avoid the monomorphism
restriction, to keep the type of |testpowfix| and |testpowfix7|
polymorphic in |repr|. \ifshort\else Instead of supplying this dummy
argument, we could have given the terms explicit polymorphic
signatures.  We however prefer for
Haskell to infer the object types for us. We could also
avoid the dummy argument by switching off the monomorphism restriction
with a compiler flag. \fi
The methods |add|, |mul|, and |leq| are quite similar, and so are
|int| and |bool|. Therefore, we often elide all but
one method of each group. The
accompanying code has the complete implementations.

@@

The functor |EX| in Fig.~\ref{fig:ocaml-examples} encodes 
our running examples |test1| and the $\mathit{power}$ function
(|testpowfix|).
The dummy argument to |test1| and |testpowfix| is an artifact of
MetaOCaml: in order for us to run a
piece of generated code, it must be polymorphic in its environment
classifier (the type variable |'c| in Figure~\ref{fig:ocaml-simple}),
so we must define our object terms as syntactic values to satisfy
the value restriction.
(Alternatively, we could have used
OCaml's rank-2 record types to maintain the necessary polymorphism.)
\ifshort\else\par\fi
Thus, we represent an object expression in
OCaml as a functor from |Symantics| to a semantic domain. This
is essentially the same as the constraint \texttt{Symantics repr =>} in the
Haskell embedding.

Comparing |Symantics| with Fig.~\ref{fig:object}
shows how to represent \emph{every} well-typed object term in the
metalanguage.
We formalize this representation by defining $H$ and~$M$,
two inductive maps from terms and types in our object language
to terms and types in Haskell and OCaml:
\begin{align}
    H(\ZZ) &= \texttt{Int} &
    M(\ZZ) &= \texttt{int} \notag\\
    H(\BB) &= \texttt{Bool} &
    M(\BB) &= \texttt{bool} \notag\\
    H(t_1 \to t_2) &= H(t_1) \mathbin{\texttt{->}} H(t_2) &
    M(t_1 \to t_2) &= M(t_1) \mathbin{\texttt{->}} M(t_2) \displaybreak[0] \\
    H(x) &= x &
    M(x) &= x \notag\\
    H(\fun{x}e) &= \texttt{lam (\textbackslash $x$ -> $H(e)$)} &
    M(\fun{x}e) &= \texttt{lam (fun $x$ -> $M(e)$)} \notag\\
    H(\fix{f}e) &= \texttt{fix (\textbackslash $x$ -> $H(e)$)} &
    M(\fix{x}e) &= \texttt{fix (fun $x$ -> $M(e)$)} \notag\\
    H(e_1 e_2) &= \texttt{app $H(e_1)$ $H(e_2)$} &
    M(e_1 e_2) &= \texttt{app $M(e_1)$ $M(e_2)$} \notag\\
    H(n) &= \texttt{int $n$} &
    M(n) &= \texttt{int $n$} \notag\\
    H(\True) &= \texttt{bool True} &
    M(\True) &= \texttt{bool true} \notag\\
    H(\False) &= \texttt{bool False} &
    M(\False) &= \texttt{bool false} \notag\\
    \!H(\cond{e}{e_1}{e_2}) &= \rlap{\texttt{if\_ $H(e)$ $H(e_1)$ $H(e_2)$}} \notag\\
    \!M(\cond{e}{e_1}{e_2}) &= \rlap{\texttt{if\_ $M(e)$ (fun () -> $M(e_1)$) (fun () -> $M(e_2)$)}} \notag\\
    H(e_1 + e_2) &= \texttt{add $H(e_1)$ $H(e_2)$} &
    M(e_1 + e_2) &= \texttt{add $M(e_1)$ $M(e_2)$} \notag\\
    H(e_1 \times e_2) &= \texttt{mul $H(e_1)$ $H(e_2)$} &
    M(e_1 \times e_2) &= \texttt{mul $M(e_1)$ $M(e_2)$} \notag\\
    H(e_1 \le e_2) &= \texttt{leq $H(e_1)$ $H(e_2)$} &
    M(e_1 \le e_2) &= \texttt{leq $M(e_1)$ $M(e_2)$}
\end{align}
These definitions assume that our object language, Haskell, and OCaml
use the same variable names~$x$ and integer literals~$n$.
If $\Gamma$ is a typing context $x_1:t_1,\dotsc,x_n:t_n$ in the object
language, then we define the metalanguage contexts
\begin{align}
    \texttt{repr\;$H(\Gamma)$}
    &= x_1:\texttt{repr}\;H(t_1),\dotsc,x_n:\texttt{repr}\;H(t_n) \text, \\
    \texttt{('c,\:$M(\Gamma)$)\;repr}
    &= x_1:\texttt{('c,\:$M(t_1)$)\;repr},\dotsc,x_n:\texttt{('c,\:$M(t_n)$)\;repr} \text.
\end{align}

@@

The first interpreter evaluates an object term to its value in the
metalanguage.  The module below is metacircular in that it
interprets each object\hyp language
operation as the corresponding metalanguage operation.
\ifshort
\begin{code3}
module R = struct type ('c,'dv) repr = 'dv (* no wrappers *)
  let int  (x:int)  = x         let bool (b:bool) = b
  let lam  f        = f         let app  e1 e2    = e1 e2
  let fix  f        = let rec self n = f self n in self
  let add  e1 e2    = e1 + e2   let mul  e1 e2    = e1 * e2
  let leq  e1 e2    = e1 <= e2
  let if_  eb et ee = if eb then et () else ee () end
\end{code3}
\else
\begin{code}
module R = struct
  type ('c,'dv) repr = 'dv (* no wrappers *)

  let int  (x:int)  = x
  let bool (b:bool) = b
  let lam  f        = f
  let app  e1 e2    = e1 e2
  let fix  f        = let rec self n = f self n in self
  let add  e1 e2    = e1 + e2
  let mul  e1 e2    = e1 * e2
  let leq  e1 e2    = e1 <= e2
  let if_  eb et ee = if eb then et () else ee ()
end
\end{code}
\fi
%
As in~\S\ref{ourapproach},
this interpreter is patently tagless, using neither a universal type nor
any pattern matching: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. To run our
examples, we instantiate the |EX| functor from~\S\ref{encoding}
with~|R|\ifshort: \texttt{module EXR = EX(R)}\fi.
\ifshort\else
\begin{code}
module EXR = EX(R)
\end{code}
\fi
Thus, |EXR.test1 ()| evaluates to the untagged boolean value |true|.
\begin{comment}
%% commenting out this Haskell in JFP version as well
In Haskell, we define
\begin{code}
newtype R a = R {unR::a}
instance Symantics R where ...
\end{code}
Although |R| looks like a tag, it is only
a |newtype|.  The types |a| and |R a| are represented differently
only at compile time, not at run time.  Pattern matching against~|R|
cannot ever fail and is assuredly compiled away.
\end{comment}
It is obvious to the compiler that
pattern matching cannot fail, because there is no
pattern matching. Evaluation can only fail to yield a value
due to interpreting |fix|.
The soundness of the object language's type system with respect to
the dynamic semantics specified by a definitional interpreter
follows from the soundness of the metalanguage's type system.
\ifshort
(The source code shows a total interpreter |L| that measures
the size of each object term.)
\fi
\ifshort
We can also generalize from~|R| to all interpreters;
these propositions follow immediately from the soundness of the
metalanguage's type system.
\fi
\begin{proposition}
If a closed object term~$e$ has type~$t$,
and the OCaml module |I| implements the signature |Symantics|,
then under the OCaml module definition
\[
    \texttt{\begin{tabular}{@{}l@{}}
    module RESULT = \\
    \quad (functor (S:Symantics) -> struct open S let term () = $M(e)$ end) \\
    \quad (I)
    \end{tabular}}
\]
evaluating the expression \texttt{RESULT.term ()} never gets stuck: it either
does not terminate or evaluates to a value of type
\texttt{('c,\:$M(t)$)\;I.repr} (polymorphic over~|'c|).
\end{proposition}
\begin{proof}
By Corollary~\ref{cor:typing} and the type soundness of (this fragment of) OCaml.
\end{proof}
\begin{corollary}
If a closed object term~$e$ has type~$t$,
then under the OCaml module definition
\[
    \texttt{\begin{tabular}{@{}l@{}}
    module RESULT = \\
    \quad (functor (S:Symantics) -> struct open S let term () = $M(e)$ end) \\
    \quad (R)
    \end{tabular}}
\]
evaluating the expression \texttt{RESULT.term ()} never gets stuck: it either
does not terminate or evaluates to a value of type~$M(t)$.
\end{corollary}
\ifshort\else
For variety, we show another interpreter, which measures the \emph{size}
(or \emph{length}, thus the mnemonic |L|)
of each object term, defined as the number of term
constructors.
\begin{code}
module L = struct
  type ('c,'dv) repr = int

@@

Although the rest of this paper continues to represent binding
using HOAS\@, our approach is compatible with de Bruijn indices.
The accompanying source file |incope-dB.ml| implements this
alternative, starting with the |Symantics| signature and the |R|
evaluator in Fig.~\ref{fig:de-Bruijn}.

@@

To turn the evaluator~|R| into a simple compiler, we bracket the
computation on values to be performed at run time, then escape the code
generation from terms to be performed at compile time.  Adding these
stage annotations yields the compiler~|C|
\ifshort in Fig.~\ref{fig:interpreter-C}(a).\else below.
\UseVerbatim{5a}
\fi
\ifshort
\begin{figure}[t]
(a) \BUseVerbatim{5a}

@@

\medskip
\caption{The tagless staged interpreter \texttt{C}}
\label{fig:interpreter-C}
\end{figure}
\fi
This is a straightforward staging of
|module R|.
This compiler produces
unoptimized code. For example, interpreting our |test1| with
\ifshort Fig.~\ref{fig:interpreter-C}(b) \else \UseVerbatim{5b}\fi
gives the code value |.<(fun|~|x_6|~|->|~|x_6)| |true>.|
of inferred type |('c,|~|bool)| |C.repr|.  Interpreting |testpowfix7|
with
\ifshort Fig.~\ref{fig:interpreter-C}(c) \else \UseVerbatim{5c}\fi
gives a code value with many apparent $\beta$- and $\eta$-redexes\ifshort,
Fig.~\ref{fig:interpreter-C}(d). \else: \UseVerbatim{5d}\fi
This compiler does not incur
any interpretive overhead: the
code produced for $\fun{x}x$ is simply |fun|~|x_6|~|->|~|x_6|\ifshort\else\
and does not
call the interpreter, unlike the recursive calls to |eval0| and
|eval| in the |L e| lines in \S\ref{tagproblem}\fi.
The resulting code obviously contains no tags and no pattern matching.
The environment classifiers here, like the tuple types in \S\ref{ourapproach},
make it a type error to run an open expression.
\ifshort
The accompanying code shows the Haskell implementation. 
\else
\begin{proposition}
If an object term~$e$ has the type~$t$
in the context $x_1:t_1,\dotsc,x_n:t_n$,
then in a MetaOCaml environment
\[
    \texttt{open C},\quad
    x_1 \mapsto \texttt{.<$y_1$>.},\enspace
    \dotsc,\enspace
    x_n \mapsto \texttt{.<$y_n$>.}
\]
where each $y_i$ is a future-stage variable of type $M(t_i)$,
the MetaOCaml term~$M(e)$ evaluates to a code value,
of type~\texttt{('c,\:$M(t)$)\;code} (polymorphic over~|'c|),
that contains no pattern-matching operations.
\end{proposition}
\begin{proof}
    By structural induction on the typing derivation of~$e$.
\end{proof}
\begin{corollary}
If a closed object term~$e$ has type~$t$,
then under the OCaml module definition
\[
    \texttt{\begin{tabular}{@{}l@{}}
    module RESULT = \\
    \quad (functor (S:Symantics) -> struct open S let term () = $M(e)$ end) \\
    \quad (C)
    \end{tabular}}
\]
the expression \texttt{RESULT.term ()} evaluates to a code value,
of type~\texttt{('c,\:$M(t)$)\;code} (polymorphic over~|'c|),
that contains no pattern-matching operations.
\end{corollary}

>>>>

% see tagless\_interp1.ml, module Tagfull for the complete code.
% If you change the code in here, please adjust the .ml file
% accordingly. Let the paper and the accompanying code be in sync.
It is straightforward to create an algebraic data type, say in OCaml%
\ifshort, Figure~\ref{fig:tag-problem}(a)\fi, to
represent object terms such as those in Figure~\ref{fig:object}.
For brevity, we elide treating integers, conditionals, and fixpoint in
this section.
\ifshort\else\UseVerbatim{2a}\fi
We represent each variable using a unary de Bruijn index.%
\footnote{We use de Bruijn indices to simplify the comparison with
\citearound{'s work}\citet{WalidICFP02}.}
For example, we represent the object term $(\fun{x}x)\True$ as
\ifshort \BUseVerbatim{test1}.\else \UseVerbatim{test1}\fi
\ifshort

@@

\fi
Following \citet{WalidICFP02},
we try to implement an interpreter function |eval0|\ifshort,
Figure~\ref{fig:tag-problem}(b)\fi. It takes
an object term such as |test1| above and gives us its value.
The first argument to |eval0| is the environment, initially empty,
which is the list of values bound to free variables in the
interpreted code.
\ifshort\else\UseVerbatim{2b}\fi
If our OCaml-like metalanguage were untyped, the code above would be 
acceptable.
The |L e| line exhibits interpretive overhead:
|eval0| traverses the function body~|e| every time (the result of
evaluating) |L e| is applied. Staging can be used to remove this
interpretive overhead \citep[\S1.1--2]{WalidICFP02}.

@@

The usual solution is to introduce a universal type \citep[\S1.3]
{WalidICFP02} containing both booleans and functions\ifshort,
Figure~\ref{fig:tag-problem}(c)\fi.
\ifshort\else\UseVerbatim{2c}\fi
We can then write a typed interpreter\ifshort,
Figure~\ref{fig:tag-problem}(d), \else\UseVerbatim{2d}\fi
whose inferred type is |u list -> exp -> u|. Now we can evaluate
\ifshort
\texttt{eval [] test1} obtaining |UB true|.
\else
\begin{code}
let test1r = eval [] test1
val test1r : u = UB true 
\end{code}
\fi
The unfortunate tag |UB| in the result reflects that |eval| is a partial
function.  
First, the pattern match |with UA f| in the line
|A (e1,e2)| is not exhaustive, so |eval| can fail if we apply a boolean,
as in the ill-typed term |A (B true, B false)|.
\ifshort\else
\begin{code}
let test2 = A (B true, B false)
let test2r = eval [] test2
Exception: Match_failure in eval
\end{code}
\fi
Second, the |lookup|
function assumes a nonempty environment, so |eval| can fail if we
evaluate an open term
\ifshort
|A (L (V (VS VZ)), B true)|.
\else
\begin{code}
let test3 = A (L (V (VS VZ)), B true)
let test3r = eval [] test3
Exception: Match_failure in lookup
\end{code}
\fi
After all, the type |exp| represents object
terms both well-typed and ill-typed, both open and closed.

@@

The evaluator above is wired directly into
functions such as |b|, |lam|, and |app|, whose names appear
free in |testf1| above.
\ifshort We \else In the rest of this paper, we \fi
explain how to abstract over these functions' definitions
and apply different folds to the same object language, so as
to process the same term using many other
interpreters: we can
\begin{itemize}
    \item evaluate the term to a value in the metalanguage;
    \item measure the length of the term;
    \item compile the term, with staging support such as in MetaOCaml;
    \item partially evaluate the term, online; and
    \item transform the term to continuation\hyp passing style (CPS),
        even call-by-name (CBN) CPS in a call-by-value (CBV) metalanguage,
        so as to isolate the evaluation
        order of the object language from that of the metalanguage.\ifshort
\footnote{Due to serious lack of space, 
we refer the reader to the accompanying code for this.}\fi
\end{itemize}
We have programmed all our interpreters and examples in OCaml (and, for staging,
\citet{metaocaml}) and standard Haskell. The complete code is
available at \url{http://okmij.org/ftp/packages/tagless-final.tar.gz}
to supplement the paper. 
\ifshort For simplicity, main examples in the paper will be in MetaOCaml;
all examples have also been implemented in Haskell.
\else Except for the basic definitions in \S\ref{encoding},
we show our examples in (Meta)OCaml even though some of our
claims are more obvious in Haskell, for consistency and because
MetaOCaml provides convenient, typed staging facilities.
\fi

@@

Figure~\ref{fig:object} shows our object language, a simply-typed
$\lambda$-calculus with fixpoint, integers, booleans, and comparison.
The language is close to \citets{xi-guarded}, without their polymorphic
lift but with more constants so as to more conveniently express Fibonacci,
factorial, and power.
In contrast to \S\ref{intro}, in the rest of the paper 
we use higher-order
abstract syntax (HOAS) \citep{miller-manipulating,pfenning-higher-order}
rather than de Bruijn indices to encode binding and
ensures that our object programs are closed.
We find HOAS to be more convenient, but we have also implemented our
approach using de Bruijn indices
(in \S\ref{S:de-Bruijn} and the accompanying source file
|incope-dB.ml|).

@@

We embed our language in (Meta)OCaml and Haskell.  In Haskell,
the functions that construct object terms are methods in a type class
|Symantics| (with a parameter |repr| of kind |* -> *|)\ifshort,
Figure~\ref{fig:symantics-haskell}(a)\fi. The class is so named
because its interface gives the syntax of the object language and its
instances give the semantics.
\ifshort\else\UseVerbatim{3a}\fi
For example, we encode the term |test1|, or $(\fun{x}x)\True$, from
\S\ref{tagproblem} above as \texttt{app (lam (\textbackslash x -> x)) (bool True)},
whose inferred type is \texttt{Symantics repr => repr Bool}.
For another example, the classical $\mathit{power}$ function is
\ifshort in Figure~\ref{fig:symantics-haskell}(b)
\else\UseVerbatim{3b}\fi
and the partial application $\fun{x} \mathit{power}\;x\;7$ is
\ifshort in Figure~\ref{fig:symantics-haskell}(c).
\else\UseVerbatim{3c}\fi
The dummy argument |()| above is to avoid the monomorphism
restriction, to keep the type of |testpowfix| and |testpowfix7|
polymorphic in |repr|. \ifshort\else Instead of supplying this dummy
argument, we could have given the terms explicit polymorphic
signatures.  We however prefer for
Haskell to infer the object types for us. We could also
avoid the dummy argument by switching off the monomorphism restriction
with a compiler flag. \fi
The methods |add|, |mul|, and |leq| are quite similar, and so are
|int| and |bool|. Therefore, we often elide all but
one method of each group. The
accompanying code has the complete implementations.

@@

The functor |EX| in Figure~\ref{fig:ocaml-examples} encodes 
our running examples |test1| and the $\mathit{power}$ function
(|testpowfix|).
The dummy argument to |test1| and |testpowfix| is an artifact of
MetaOCaml: in order for us to run a
piece of generated code, it must be polymorphic in its environment
classifier (the type variable |'c| in Figure~\ref{fig:ocaml-simple}),
so we must define our object terms as syntactic values to satisfy
the value restriction.
(Alternatively, we could have used
OCaml's rank-2 record types to maintain the necessary polymorphism.)
\ifshort\else\par\fi
Thus, we represent an object expression in
OCaml as a functor from |Symantics| to a semantic domain. This
is essentially the same as the constraint \texttt{Symantics repr =>} in the
Haskell embedding.

Comparing |Symantics| with Figure~\ref{fig:object}
shows how to represent \emph{every} well-typed object term in the
metalanguage.
We formalize this representation by defining $H$ and~$M$,
two inductive maps from terms and types in our object language
to terms and types in Haskell and OCaml:
\begin{align}
    H(\ZZ) &= \texttt{Int} &
    M(\ZZ) &= \texttt{int} \notag\\
    H(\BB) &= \texttt{Bool} &
    M(\BB) &= \texttt{bool} \notag\\
    H(t_1 \to t_2) &= H(t_1) \mathbin{\texttt{->}} H(t_2) &
    M(t_1 \to t_2) &= M(t_1) \mathbin{\texttt{->}} M(t_2) \displaybreak[0] \\
    H(x) &= x &
    M(x) &= x \notag\\
    H(\fun{x}e) &= \texttt{lam (\textbackslash $x$ -> $H(e)$)} &
    M(\fun{x}e) &= \texttt{lam (fun $x$ -> $M(e)$)} \notag\\
    H(\fix{f}e) &= \texttt{fix (\textbackslash $x$ -> $H(e)$)} &
    M(\fix{x}e) &= \texttt{fix (fun $x$ -> $M(e)$)} \notag\\
    H(e_1 e_2) &= \texttt{app $H(e_1)$ $H(e_2)$} &
    M(e_1 e_2) &= \texttt{app $M(e_1)$ $M(e_2)$} \notag\\
    H(n) &= \texttt{int $n$} &
    M(n) &= \texttt{int $n$} \notag\\
    H(\True) &= \texttt{bool True} &
    M(\True) &= \texttt{bool true} \notag\\
    H(\False) &= \texttt{bool False} &
    M(\False) &= \texttt{bool false} \notag\\
    \!H(\cond{e}{e_1}{e_2}) &= \rlap{\texttt{if\_ $H(e)$ $H(e_1)$ $H(e_2)$}} \notag\\
    \!M(\cond{e}{e_1}{e_2}) &= \rlap{\texttt{if\_ $M(e)$ (fun () -> $M(e_1)$) (fun () -> $M(e_2)$)}} \notag\\
    H(e_1 + e_2) &= \texttt{add $H(e_1)$ $H(e_2)$} &
    M(e_1 + e_2) &= \texttt{add $M(e_1)$ $M(e_2)$} \notag\\
    H(e_1 \times e_2) &= \texttt{mul $H(e_1)$ $H(e_2)$} &
    M(e_1 \times e_2) &= \texttt{mul $M(e_1)$ $M(e_2)$} \notag\\
    H(e_1 \le e_2) &= \texttt{leq $H(e_1)$ $H(e_2)$} &
    M(e_1 \le e_2) &= \texttt{leq $M(e_1)$ $M(e_2)$}
\end{align}
These definitions assume that our object language, Haskell, and OCaml
use the same variable names~$x$ and integer literals~$n$.
If $\Gamma$ is a typing context $x_1:t_1,\dotsc,x_n:t_n$ in the object
language, then we define the metalanguage contexts
\begin{align}
    \texttt{repr\;$H(\Gamma)$}
    &= x_1:\texttt{repr}\;H(t_1),\dotsc,x_n:\texttt{repr}\;H(t_n) \text, \\
    \texttt{('c,\:$M(\Gamma)$)\;repr}
    &= x_1:\texttt{('c,\:$M(t_1)$)\;repr},\dotsc,x_n:\texttt{('c,\:$M(t_n)$)\;repr} \text.
\end{align}

@@

The first interpreter evaluates an object term to its value in the
metalanguage.  The module |R| below is metacircular in that it
\emph{runs} each object\hyp language
operation by executing the corresponding metalanguage operation.
\ifshort
\begin{code3}
module R = struct type ('c,'dv) repr = 'dv (* no wrappers *)
  let int  (x:int)  = x         let bool (b:bool) = b
  let lam  f        = f         let app  e1 e2    = e1 e2
  let fix  f        = let rec self n = f self n in self
  let add  e1 e2    = e1 + e2   let mul  e1 e2    = e1 * e2
  let leq  e1 e2    = e1 <= e2
  let if_  eb et ee = if eb then et () else ee () end
\end{code3}
\else
\begin{code}
module R = struct
  type ('c,'dv) repr = 'dv (* no wrappers *)

  let int  (x:int)  = x
  let bool (b:bool) = b
  let lam  f        = f
  let app  e1 e2    = e1 e2
  let fix  f        = let rec self n = f self n in self
  let add  e1 e2    = e1 + e2
  let mul  e1 e2    = e1 * e2
  let leq  e1 e2    = e1 <= e2
  let if_  eb et ee = if eb then et () else ee ()
end
\end{code}
\fi
%
As in~\S\ref{ourapproach},
this interpreter is patently tagless, using neither a universal type nor
any pattern matching: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. To run our
examples, we instantiate the |EX| functor from~\S\ref{encoding}
with~|R|\ifshort: \texttt{module EXR = EX(R)}\fi.
\ifshort\else
\begin{code}
module EXR = EX(R)
\end{code}
\fi
Thus, |EXR.test1 ()| evaluates to the untagged boolean value |true|.
\begin{comment}
%% commenting out this Haskell in JFP version as well
In Haskell, we define
\begin{code}
newtype R a = R {unR::a}
instance Symantics R where ...
\end{code}
Although |R| looks like a tag, it is only
a |newtype|.  The types |a| and |R a| are represented differently
only at compile time, not at run time.  Pattern matching against~|R|
cannot ever fail and is assuredly compiled away.
\end{comment}
It is obvious to the compiler that
pattern matching cannot fail, because there is no
pattern matching. Evaluation can only fail to yield a value
due to interpreting |fix|.
The soundness of the object language's type system with respect to
the dynamic semantics specified by a definitional interpreter
follows from the soundness of the metalanguage's type system.
\ifshort
(The source code shows a total interpreter |L| that measures
the length of each object term.)
\fi
\ifshort
We can also generalize from~|R| to all interpreters;
these propositions follow immediately from the soundness of the
metalanguage's type system.
\fi
\begin{proposition}
If a closed object term~$e$ has type~$t$,
and the OCaml module |I| implements the signature |Symantics|,
then under the OCaml module definition
\[
    \texttt{\begin{tabular}{@{}l@{}}
    module RESULT = \\
    \quad (functor (S:Symantics) -> struct open S let term () = $M(e)$ end) \\
    \quad (I)
    \end{tabular}}
\]
evaluating the expression \texttt{RESULT.term ()} never gets stuck: it either
does not terminate or evaluates to a value of type
\texttt{('c,\:$M(t)$)\;I.repr} (polymorphic over~|'c|).
\end{proposition}
\begin{proof}
By Corollary~\ref{cor:typing} and the type soundness of (this fragment of) OCaml.
\end{proof}
\begin{corollary}
If a closed object term~$e$ has type~$t$,
then under the OCaml module definition
\[
    \texttt{\begin{tabular}{@{}l@{}}
    module RESULT = \\
    \quad (functor (S:Symantics) -> struct open S let term () = $M(e)$ end) \\
    \quad (R)
    \end{tabular}}
\]
evaluating the expression \texttt{RESULT.term ()} never gets stuck: it either
does not terminate or evaluates to a value of type~$M(t)$.
\end{corollary}
\ifshort\else
For variety, we show another interpreter~|L|, which measures the
\emph{length}
of each object term, defined as the number of term
constructors.
\begin{code}
module L = struct
  type ('c,'dv) repr = int

@@

Although the rest of this paper continues to represent binding
using HOAS\@, our approach is compatible with de Bruijn indices.
The accompanying source file |incope-dB.ml| implements this
alternative, starting with the |Symantics| signature and the |R|
evaluator in Figure~\ref{fig:de-Bruijn}.

@@

To turn the evaluator~|R| into a simple compiler, we bracket the
computation on values to be performed at run time, then escape the code
generation from terms to be performed at compile time.  Adding these
stage annotations yields the compiler~|C|
\ifshort in Figure~\ref{fig:interpreter-C}(a).\else below.
\UseVerbatim{5a}
\fi
\ifshort
\begin{figure}[t]
(a) \BUseVerbatim{5a}

@@

\medskip
\caption{The tagless staged interpreter \texttt{C}}
\label{fig:interpreter-C}
\end{figure}
\fi
This is a straightforward staging of
|module R|.
This compiler produces
unoptimized code. For example, interpreting our |test1| with
\ifshort Figure~\ref{fig:interpreter-C}(b) \else \UseVerbatim{5b}\fi
gives the code value |.<(fun|~|x_6|~|->|~|x_6)| |true>.|
of inferred type |('c,|~|bool)| |C.repr|.  Interpreting |testpowfix7|
with
\ifshort Figure~\ref{fig:interpreter-C}(c) \else \UseVerbatim{5c}\fi
gives a code value with many apparent $\beta$- and $\eta$-redexes\ifshort,
Figure~\ref{fig:interpreter-C}(d). \else: \UseVerbatim{5d}\fi
This compiler does not incur
any interpretive overhead: the
code produced for $\fun{x}x$ is simply |fun|~|x_6|~|->|~|x_6|\ifshort\else\
and does not
call the interpreter, unlike the recursive calls to |eval0| and
|eval| in the |L e| lines in \S\ref{tagproblem}\fi.
The resulting code obviously contains no tags and no pattern matching.
The environment classifiers here, like the tuple types in \S\ref{ourapproach},
make it a type error to run an open expression.
\ifshort
The accompanying code shows the Haskell implementation. 
\else
\begin{proposition}
If an object term~$e$ has the type~$t$
in the context $x_1:t_1,\dotsc,x_n:t_n$,
then in a MetaOCaml environment
\[
    \texttt{open C},\quad
    x_1 \mapsto \texttt{.<$y_1$>.},\enspace
    \dotsc,\enspace
    x_n \mapsto \texttt{.<$y_n$>.}
\]
where each $y_i$ is a future-stage variable of type $M(t_i)$,
the MetaOCaml term~$M(e)$ evaluates to a code value,
of type~\texttt{('c,\:$M(t)$)\;code} (polymorphic over~|'c|),
that contains no pattern-matching operations.
\end{proposition}
\begin{proof}
    By structural induction on the typing derivation of~$e$.
\end{proof}
\begin{corollary}
If a closed object term~$e$ has type~$t$,
then under the OCaml module definition
\[
    \texttt{\begin{tabular}{@{}l@{}}
    module RESULT = \\
    \quad (functor (S:Symantics) -> struct open S let term () = $M(e)$ end) \\
    \quad (C)
    \end{tabular}}
\]
the expression \texttt{RESULT.term ()} evaluates to a code value,
of type~\texttt{('c,\:$M(t)$)\;code} (polymorphic over~|'c|),
that contains no pattern-matching operations.
\end{corollary}
